---
title: "Sampling"
---

Follow along with the demo to learn how to sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load 1,000 rows of data with the ages of people from 18-80

```{r}
set.seed(1234)

population <- tibble(
  age = sample (18:80, 1000, replace = TRUE)
)
```

### Task 3: Sample 100 rows of data from the dataset

```{r}
sample100 <- population |>
  sample_n(100)
```

### Task 4: Calculate the mean age and standard deviation of the sample

```{r}
sample100 |>
  summarize(
    mean_age = mean(age),
    sd_age = sd(age)
  )
```

### Task 5: Calculate the mean age and standard deviation of the population

```{r}
population |>
  summarize(
    mean_age = mean(age),
    sd_age = sd(age)
  )
```

### Task 6: Compare the sample and population statistics

Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?
The sample population mean and standard deviation is only slightly higher than the population mean and standard deviation. For example, the sample mean age is 49.82 and the population mean age is 49.06. Since aren't measured by decimals, the fact that both sample and population mean ages would be 49 means they are extremely similar, practically the same.

The same happens for standard devitaitons. The sample standard deviation is 17.88, whereas the population standard deviation is 17.77. 

The sample mean and standard deviation ages are close to the population's mean and standard deviation ages.


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
