---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")

The crimes that occur the least are Breaking & Entering, Homicide and Residential Robbery. Accidents occur at the highest rate. Theft from Auto, Auto Stolen and Accident with Impound also occur higher than most other categories. Accident incidences have risen, with the lowest incident rate being 429 in March and the highest being 529 in May. 
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

Accident rates occur more than other categories. No category is specifically more important than others because any crime affecting the local community should be reported on, as they all the local community. However, the amount of accidents make me wonder where these incidents occur. The accident category could also be specified as auto accidents for clarification, but detail doesn't necessarily make summarizing the data different. Since accidents occur more than the other categories, as a reporter, it's worth investigating what is causing them-- traffic violations? Traffic light timing? Pedestrians? Driving under the influence? Also, is there correlation between the high rates of accidents with auto theft, accident with impound and auto stolen? 

### Task 4: Calculating Averages

Add up the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime using the total number of months, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime_totals <- pg_crime |> mutate(total_incidents = Jan + Feb + Mar + Apr + May + Jun + Jul + Aug + Sep + Oct + Nov + Dec)

pg_crime_averages <- pg_crime_totals |> mutate(`average_incidents` = `total_incidents` / 12)
```

### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

On average, the categories with the highest number of incidences per month are accident, auto stolen, theft from auto, accident with impound and theft. The lowest incidences on average are residential robberty (N/A), homicide, breaking and entering (other) and sex offense. Based on the data, auto-related crimes happen at a higher rate than robberies, sexual assault, assault with weapon or breaking and entering. I wonder if the auto-related crimes have any intersection and if there are any locations where the crimes occur more. This also makes me question PGPD's efforts to curb auto-related crimes if there is a specifc location where these occur more. 


### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime_averages |>
  mutate(
    june_change = (`Jun` - average_incidents) / average_incidents * 100,
    july_change = (`Jul` - average_incidents) / average_incidents * 100,
    august_change = (`Aug` - average_incidents) / average_incidents * 100
)
```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

Assault shooting crimes show an increase in percent change when comparing the percent change of June (-4.36%) to August (34.04%). Other crimes are relatively similar in percent change, such as accident, which saw its peak in percent change in August at 3.25%, but had a -4.02% change in August and -3.81% in June. The auto-related crimes did not jump as high in percent change compared to assault shooting. In terms of crime being more likely to occur in the summer months, this can be backed by July having the highest percentage changes (e.g., vehicle robbery was at 64.55% in July and 24.41% in June; sex offense at 22.45% in July and 14.29% in June, but -18.37% in August; commercial robbery had a percent change at 43.80% in August, but was at -0.83% in July and -15.70% in June). 

### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Auto-related crimes in Prince George's County show an uptick in occurrences compared to other crimes on average, based on an analysis from the Prince George’s County Police Department's crime reports for 2024. 

Accidents occur the highest at 481.33 on average, followed by stolen auto vehicles at 399.17, theft from an auto vehicle at 391.92 and accident with impound at 320.75. Theft occurs at an 284.08 average, assault at 70.92, robberies listed under 'other' at 54.92 and commercial breaking and entering at 44.67 in comparison to auto-related crimes. The frequency of auto crimes signals a public-safety concern for PGC residents.
  

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
