---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |> mutate(violent_rate_2019 = violent2019/pop2019 * 1000, property_rate_2019 = property2019/pop2019 * 1000, violent_rate_2020 = violent2020/pop2020 * 1000, property_rate_2020 = property2019/pop2020 * 1000)

```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

The cities with the highest property crime rates were (in order) Ocean City (103.67), Oakland (73.28) and Elkton (68.11) in 2019. For 2020 property crime rates, the cities with the highest rates remain the same. However, the rates changed-- Ocean City rose to 103.96, Oakland rose to 73.64 and Elkton decreased to 68.04. 

The cities with the highest violent crime rates in 2019 were Baltimore (18.34), Ocean City (13.10) and Williamsport (12.92). In 2020, the cities with the highest violent crime rates were Ocean City (23.97), Baltimore (15.97) and Cambridge (15.69). By 2020, Ocean City surpassed Baltimore in highest violent crime rates. Williamsport was surpassed by Cambridge.

Urban or populated areas are more likely to have high crime rates in Maryland. For example, Ocean City appeared on both property and violent crimes for 2019 and 2020. Property crimes strikingly occurred at a higher rate both years. A story angle I could take would be debunking the idea that Baltimore is known for crime when, looking at property crimes, Baltimore ranked 14th in 2019 and 10th in 2020. This story could also tie in violent crime rates by comparing Baltimore's rates to other high-ranking violent crime rate cities, which have rates relatively close to Baltimore's. The story could compare Ocean City's violent crime rates Baltimore's in 2020. 

(Side note, Williamsport surprised me because I live 5 minutes away from it.)


### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates |> 
  mutate(
    violent_crimes_pct_change = (violent_rate_2020 - violent_rate_2019) / violent_rate_2019 * 100,
    property_crime_pct_change = (property_rate_2020 - property_rate_2019) / property_rate_2019 * 100
  )
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

Something that stood out to me about the violent_crimes_pct_change was that Chevy Chase Sector V, Chevy Chase Sector III, Charlestown, Upper Marlboro, Friendsville and Laytonsville had "Inf" instead of values. A lot of other cities had "NaN" for their values. This makes inferring the data hard because there's no concrete values to analyze these cities with. I would call the cities with "Inf" and "NaN" outliers for that reason. 

The cities with the highest violent crime percentage changes were Edmonston (895.32%), Poolesville (607.15%) and Pocomoke City (300.98%). For violent crime percent changes, the cities that saw the lowest all had -100.00%: Cecilton, Friendship Heights, Myersville, New Windsor, North Chevy Chase, Pittsville, Port Deposit, Queenstown, Ridgely, Rock Hall, Rosemont and Sharpsburg. However, I would call these cities outliers because their percentage changes could be influenced by their 2020 property rates, which 0. To me, the outliers are more interesting.

The lowest property crime percentage changes were Clear Spring (-24.43%), Sharpsburg (-17.47%) and Laytonsville (-16.45%). Clear Spring, Sharpsburg and Myersville (though it says "NaN") are all in Washington County. As a reporter, I would look at the efforts Washington County police departments have made towards curbing property crime from 2019 to 2020. The highest property crime percentage changes were Kitzmiller (29.49%), Garrett Park (20.16%) and Somerset (16.79%). Garrett Park and Somerset are in Montgomery County, so it would be worth looking at Montgomery County police departments' efforts to handle property crimes in comparison to Washington County's.  

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Ocean City surpassed Baltimore in violent crime rates in 2020, placing Ocean City as Maryland's highest ranking city for violent crimes.

Baltimore had 18.34 violent crimes per 1,000 people, whereas Ocean City had 13.10 violent crimes per 1,000 people in 2019. However, the rates shifted in 2020 when Ocean City presented 23.97 violent crimes per 1,000 people and Baltimore saw 15.97 per 1,000 people. There was a 82.89% change in violent crimes between these years for Ocean City, giving Ocean City a staggering increase. 


### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

### Task 10

- Using the 2020 violent crime rate, calculate the percentiles using `ntile()`. Save the results to a new dataframe called `md_cities_percentiles`. How many cities are above the 90th percentile?

```{r}
md_cities_percentiles <- md_cities_rates |> mutate(percentile = ntile(violent_rate_2020, 100))
```

10 cities are above the 90th percentile.

- Calculate the Interquartile Range of the 2020 violent crime rate using `summarize()` and R's `IQR()` function. The syntax is similar to how `mean()` works.

```{r}
md_cities_percentiles |> summarize(violent_crimes_IQR = IQR(violent_rate_2020))
```
- Do the results change your best story idea from the previous assignment? Why or why not?
These results don't change my best story idea. It supports my angle that Ocean City and Baltimore presented the most violent crimes in 2020 for Maryland because both cities sit higher outside of the IQR.
